---
title: "Monthly County Level Excess Mortality in the United States, 2020-2022"
output: github_document
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  dev.args = list(png = list(type = "cairo")), 
  dpi = 96,
  fig.path = "README_files/"
)

options(
  scipen = 999,
  digits = 3,
  knitr.kable.NA = "n/a"
)

```

**Table of Contents**

1. Introduction
2. Overview of Repository
3. Available Data Files
4. Data Dictionaries and Descriptive Statistics


## Introduction

This repo includes data and code required to replicate the results in the paper ["Excess all-cause mortality across counties in the United States, March 2020 to December 2021" [Link to article in Science Advances]](https://www.science.org/doi/10.1126/sciadv.adf9742) and ["Differences Between Reported COVID-19 Deaths and Estimated Excess Deaths in Counties Across the United States, March 2020 to February 2022" [Link to pre-print]](https://www.medrxiv.org/content/10.1101/2023.01.16.23284633v1).

The aim of this project is to estimate monthly deaths at the county level for the United States for the period March 2020 - August 2022 in the counterfactual scenario of no COVID-19 pandemic. We do so with a Bayesian hierarchical model with a flexible time component and a spatial component. To deal with suppression of death counts < 10, we employ a set of state-year censored Poisson models to estimate the censored death counts. We then adjust the total number of imputed deaths making sure that the sum of non-suppressed and imputed deaths equal the state yearly death count (very unlikely to be suppressed).

## Overview of Repository

### General repository structure

The repository is organized in several directories:

- `Python`: Contains the Python code used to generate monthly population estimates.
- `R`: Contains the R code used to prepare the data, train the models, assess the models' perfomance, produce the estimates of excess mortality, and produce the figures and tables for the paper.
- `data/input`: Contains all the data needed to train the models and produce estimates of excess deaths. To reconcile inconsistencies across the various data sources we are using, and to accomodate further analysis with a longer backward time frame (i.e. we trained our final models on data for 2015-2019 but initially considered a wider window), we harmonized county FIPS code following the schema in the FIPSFixes.csv file.
- `data/output`: Will contain all the data that will be generated as a product of the project. We already included different sets of estimates compliant with the CDC Wonder user agreement.
- `figures`: contains all the figures and tables appearing in the paper.

### Main programs

The project consists of several RMarkdown and R files:

- `cleanMortalityData.Rmd`: Cleans the all-cause mortality data.
- `cleanCOVIDData.Rmd`: Cleans the data for COVID mortality.
- `modelFitMonthly.Rmd`: Fit the monthly level model for all-cause deaths.
- `modelSummary.Rmd`: Extracts the model's parameters and hyperparameters and creates summary tables and graphs.
- `crossValidationMonthly.Rmd`: Runs cross-validation on the model and produces a set of estimates to be used to evaluate the model.
- `modelEvaluationMonthly.Rmd`: Evaluates model performance using the cross-validation data.
- `createSimsDF.Rmd`: Creates a dataframe with county-month samples from the model's posterior distribution for death counts. This dataframe is used to construct estimates of death counts with posterior intervals at different levels of aggregation.
- `estimates*.Rmd`: These files (estimatesMonthly, estimatesPandemicYears, estimatesWaves, estimatesStates, estimatesStateMonths, and estimatesMonthlyTotals) produce estimates of expected death counts, excess death counts, and relative excess at different levels of temporal and geographical aggregation.
- `summaryTable.Rmd`: Produces a metro-division level table summarizing the key results by pandemic year.
- `summaryTableFinal.Rmd`: Produces an alternative metro-division level table summarizing the key results for the entire period.
- `scatterPlots`: Produces scatter plots of excess mortality in year 1 (March 2020 - February 2021) and year 2 (March 2021 - February 2022) and of excess mortality against COVID-19 mortality.
- `timeBarsGraphDivision.Rmd`: Produces a graph showing how relative excess evolved during the pandemic for each division, separating large metro areas and other areas.
- `plotsForSubmission.Rmd`: Creates the geofacet plot and the heatmap plot of count-level excess mortality.
- `map_by_period_relative_exc_to_eugenio.R`: Creates maps of relative excess.
- countyPlots: Produces a set of county-level plots of observed vs. expected deaths.

The files are intended to be run in the order in which they appear on this list. Detailed instructions on the steps we used to download data from the CDC WONDER platform are given in `data/output/CDCWONDERDataDetails.md`.

Aside from the R files within the R folder, we also provide the Python code we used to create monthly population estimates at the county level by interpolating yearly intercensal estimates from the Census Bureau. The Python code is in the form of a Jupyter Notebook. The final population estimates are stored in the output folder.

All estimates can be produced for a combination of all-cause (AC) or natural-cause (NC) mortality data and COVID as Underlying Cause of Death (UCD) or COVID as a contributing cause of death (MCD).

## Available Data Files

```{r libraries}
library(tidyverse)
library(vtable)
library(labelled)

source(here::here("R", "twee.R"))
```

Public versions of our estimates are available in the `data/output` directory.

Datasets are sorted in sub-directories based on the underlying data used to produce them as follows.

Expected mortality was modeled as either:

- **AC**: all-cause mortality
- **NC**: natural-cause mortality, excluding external causes of death

Deaths attributed to COVID-19 were defined as either:

- **UCD**: underlying cause of death (COVID-19 was listed as the single underlying cause of death on a death certificate)
- **MCD**: multiple causes of death (COVID-19 was listed as one of multiple causes of death on a death certificate)

The following datasets are available:

```{r tree, comment = NA}
twee(here::here("data", "output", "estimates"))
```

## Data Dictionaries and Descriptive Statistics

The following tables contain variable names, descriptions, and descriptive statistics for each of the datasets in this repository.

```{r variable descriptions}
variableDescriptions <-
  list(
    FIPSCode = glue::glue(
      "5-digit Federal Information Processing Standards (FIPS) code that uniquely identifies United States counties.
      First 2 digits represent state, final 3 digits represent county within that state.
      Includes leading zeroes.
      Place of residence of the decedents."
      ),
    countyName = glue::glue(
      "Name of the county or county-like area.
      Place of residence of the decedents."
    ),
    state = glue::glue(
      "Full state name.
      Place of residence of the decedents."
    ),
    stateFIPS = glue::glue(
      "2-digit code that uniquely identifies United States states, territories, and outlying regions.
      Includes leading zeroes.
      Place of residence of the decedents."
    ),
    census_region = glue::glue(
      "Name of the US Census Region.
      Census Divisions are nested within Census Regions.
      Place of residence of the decedents."
    ),
    census_division = glue::glue(
      "Name of the US Census Division.
      States are nested within Census Divisions, which are further nested within Census Regions.
      Place of residence of the decedents."
    ),
    metroCat = glue::glue(
      "2013 NCHS Urban-Rural Classification Scheme for Counties"
    ),
    year = glue::glue(
      "Calendar year."
    ),
    month = glue::glue(
      "Calendar month."
    ),
    monthYear = glue::glue(
      "Date of the first day of the month represented by the data.
      Of the format YYYY-MM-DD.
      DD always equals 01."
    ),
    period = glue::glue(
      "Month range that the data represent."
    ),
    pandemicYear = glue::glue(
      "12-month ranges beginning in March 2020, each representing one year of the COVID-19 pandemic in the United States.
      Note that the most recent pandemic year be be incomplete due to data lags.
      If this is the case, the pandemic year will include fewer than 12 months, as indicated in this field."
    ),
    suppressed = glue::glue(
      "Flag indicating whether output has been suppressed.
      The CDC WONDER data use agreement prohibits presenting or publishing death counts of 9 or fewer, or death rates based on counts of 9 or fewer.
      
      1 = suppression applied, 0 = not suppressed."
    ),
    COVIDDeathsUCD = glue::glue(
      "Count of deaths reported in the the National Vital Statistics System (NVSS) with COVID-19 as the underlying cause of death (ICD-10 diagnosis code = U07.1).
      Obtained via CDC WONDER.
      Counts of 9 or fewer are suppressed."
    ),
    COVIDDeathsMCD = glue::glue(
      "Count of deaths reported in the the National Vital Statistics System (NVSS) with COVID-19 as the contributing cause of death (ICD-10 diagnosis code = U07.1).
      Obtained via CDC WONDER.
      Counts of 9 or fewer are suppressed."
    ),
    deaths = glue::glue(
      "Count of deaths reported in the National Vital Statistics System (NVSS) obtained via CDC WONDER.
      Counts of 9 or fewer are suppressed."
    ),
    imputedDeaths = glue::glue(
      "Count of deaths with values imputed for counts of 9 or fewer."
    ),
    status = glue::glue(
      "Flag indicating whether a value of imputedDeaths has been imputed.
      
      0 = imputed value, 1 = original value from CDC WONDER"
    ),
    expDeathsMean = glue::glue(
      "Mean model-predicted expected deaths count."
    ),
    expDeathsLow = glue::glue(
      "Lower 5th percentile of the posterior interval for model-predicted expected deaths count."
    ),
    expDeathsMed = glue::glue(
      "Median model-predicted expected deaths count."
    ),
    expDeathsUp = glue::glue(
      "Upper 95th percentile of the posterior interval for model-predicted expected deaths count."
    ),
    excDeathsMean = glue::glue(
      "Mean estimated excess deaths, calculated as (deaths - expDeaths)."
    ),
    excDeathsLow = glue::glue(
      "Lower 5th percentile of estimated excess deaths, calculated as (deaths - expDeaths)."
    ),
    excDeathsMed = glue::glue(
      "Median estimated excess deaths, calculated as (deaths - expDeaths)."
    ),
    excDeathsUp = glue::glue(
      "Upper 95th percentile of estimated excess deaths, calculated as (deaths - expDeaths)."
    ),
    probPositiveExcess = glue::glue(
      "Probability that excess deaths exceeds zero."
    ),
    relExcDeathsMean = glue::glue(
      "Mean relative excess deaths, calculated as (excDeaths / expDeaths)."
    ),
    relExcDeathsLow = glue::glue(
      "Lower 5th percentile of relative excess deaths, calculated as (excDeaths / expDeaths)."
    ),
    relExcDeathsMed = glue::glue(
      "Median relative excess deaths, calculated as (excDeaths / expDeaths)."
    ),
    relExcDeathsUp = glue::glue(
      "Upper 95th percentile of relative excess deaths, calculated as (excDeaths / expDeaths)."
    ),
    ratioMeanUCD = glue::glue(
      "Mean ratio of excess deaths to COVID-19 deaths (as underlying cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioLowUCD = glue::glue(
      "Lower 5th percentile of ratio of excess deaths to COVID-19 deaths (as underlying cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioMedUCD = glue::glue(
      "Median ratio of excess deaths to COVID-19 deaths (as underlying cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioUpUCD = glue::glue(
      "Upper 95th percentile of ratio of excess deaths to COVID-19 deaths (as underlying cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    probPositiveNCExcessUCD = glue::glue(
      "Probability that excess deaths not assigned to COVID-19 (as underlying cause), exceeds zero."
    ),
    ratioMeanMCD = glue::glue(
      "Mean ratio of excess deaths to COVID-19 deaths (as contributing cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioLowMCD = glue::glue(
      "Lower 5th percentile of ratio of excess deaths to COVID-19 deaths (as contributing cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioMedMCD = glue::glue(
      "Median ratio of excess deaths to COVID-19 deaths (as contributing cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    ratioUpMCD = glue::glue(
      "Upper 95th percentile of ratio of excess deaths to COVID-19 deaths (as contributing cause), calculated as (excDeaths / COVIDDeaths)."
    ),
    probPositiveNCExcessMCD = glue::glue(
      "Probability that excess deaths not assigned to COVID-19 (as contributing cause), exceeds zero."
    )
  )
```

```{r data dictionary function}
autoDictionary <- function(.file_path, 
                           .labels = NA, 
                           parent_dir = FALSE,
                           ...) {

  .title <- str_remove(basename(.file_path), ".csv")
  
  if(parent_dir) {
    dir <- str_extract(dirname(.file_path), "[[:alpha:]]+$")
    
    .title <- glue::glue("{dir}/{.title}")
  }
  
  message(glue::glue("Processing { .title }"))
  
    .data <- readr::read_csv(
            file.path(.file_path),
            lazy = TRUE
          )
  
  datLabelled <- .data %>% 
    labelled::set_variable_labels(
      ., 
      .labels = .labels, 
      .strict = FALSE,
      ...
    )  
  
  rm(.data)
  
  datLabelled %>% 
  vtable::vtable(
    .,
    out = "kable",
    data.title = .title,
    lush = TRUE,
    desc = "",
    ...
  ) %>% 
    kableExtra::kable_styling(
      font_size = 12
    ) %>% 
    print()
  
  rm(datLabelled)
  
}
```

```{r print dictionary, results='asis'}
fileNamesAC <- list.files(
  here::here("data", "output", "estimates",'AC'),
  pattern = ".csv",
  full.names = TRUE,
  recursive = TRUE
)

fileNamesNC <- list.files(
  here::here("data", "output", "estimates",'NC'),
  pattern = ".csv",
  full.names = TRUE,
  recursive = TRUE
)

fileNames <- c(fileNamesAC,fileNamesNC)

walk(
    fileNames,
    ~ autoDictionary(
          .x,
          .labels = variableDescriptions,
          parent_dir = TRUE
      )
  )
```
