---
title: "Model Summary"
author: "Eugenio Paglino"
output: html_document
---

```{r, include=F}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

```{r}
# Loading necessary packages
library(INLA)
library(lubridate)
library(glue)
library(here)
library(ggrepel)
library(scico)
library(gt)
library(tidyverse)
```

```{r}
rm(list=ls())

i_am('R/modelSummary.Rmd')

inDir <- here('data','input')
outDir <- here('data','output')
figuresDir <- here('figures')

dataType <- 'NC'
dataInputFolder <- here(outDir,glue('cleanedMortality{dataType}Data'))
dataOutputFolder <- here(outDir,'modelsAndSimulations',glue('model{dataType}'))
validationFolder <- here(outDir,'validation',glue('validation{dataType}'))

set.seed(42)
```

```{r}
load(here(dataOutputFolder,'INLAModel.RData'))
```

```{r}
# Construct data frame of state abbreviations + divisions, plus DC
census_divisions <- tibble(
  division = unique(state.division)
) 

census_divisions <- census_divisions %>%
  group_by(division) %>%
  mutate(divisionID = cur_group_id()) %>%
  ungroup()
```

We extract the summary for the fixed parameters

```{r}
fixedParams <- INLAModel %$%
  summary.fixed %>% 
  rownames_to_column('name') %>%
  mutate(name=str_replace_all(name,'as.factor',''),
         name=str_replace_all(name,'\\(',''),
         name=str_replace_all(name,'\\)','')) %>%
  select(name,mean,sd,`0.025quant`,`0.5quant`,`0.975quant`)
```

We extract the summary for the random parameters

```{r}
randomParams <- map2_dfr(INLAModel$summary.random,
                         names(INLAModel$summary.random),
                         ~ .x %>%
                           mutate(paramName = .y)) %>%
  filter(paramName != 'FIPSID') %>%
  mutate(name=str_c(paramName,ID,sep = '')) %>%
  select(name,mean,sd,`0.025quant`,`0.5quant`,`0.975quant`)
```

Finally, we extract the summary for the hyper parameters

```{r}
hyperParams <- INLAModel %$%
  summary.hyperpar %>% 
  rownames_to_column('name') %>%
  select(name,mean,sd,`0.025quant`,`0.5quant`,`0.975quant`)

hyperParams <- hyperParams %>%
  add_row(filter(fixedParams,name=='timeID'))
```

## Visualize fixed parameters

Here we wish to show the Census Division specific month effects (capturing the seasonal trends in mortality). The complicated part is to convert the interaction parameters into the quantities of interest. It is just a matter of summing the baseline parameters and the interaction parameters in the correct way and then add the intercept.

```{r}
fixedParams <- fixedParams %>%
  mutate(divisionID=str_match(name,'divisionID(\\d+)')[,2],
         divisionID=if_else(is.na(divisionID),1L,as.integer(divisionID)),
         month=str_match(name,'month(\\d+)')[,2],
         month=if_else(is.na(month),1L,as.integer(month)))
```

```{r}
seasonalityParams <- fixedParams %>%
  filter(!(name %in% c('Intercept','timeID')))

seasonalityParams <- seasonalityParams %>%
  mutate(intercept = fixedParams %>% filter(name=='Intercept') %>% pull(mean),
         mean = mean + intercept,
         mean = mean - mean(mean),
         `0.025quant` = mean - 1.96*sd,
         `0.975quant` = mean + 1.96*sd)

seasonalityParams <- seasonalityParams %>%
  mutate(monthDate = if_else(month<=8,make_date(2001,month,1),make_date(2000,month,1)))
```

We are now ready for a visualization.

```{r}
seasonalityParamsPlot <- seasonalityParams %>%
  ggplot() +
  geom_ribbon(mapping=aes(x=monthDate,ymin=`0.025quant`,ymax= `0.975quant`),alpha=0.5) +
  geom_line(mapping=aes(x=monthDate,y=mean)) +
  scale_x_date(date_breaks = '1 month', date_labels = '%b') +
  labs(x='',
       y='Deviation from Mean') +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin = unit(c(1,10,1,1), "lines")) +
  guides(color='none')

seasonalityParamsPlot
```

```{r}
pdf(here(validationFolder,'seasonalityParamsPlot.pdf'), width = 8, height = 10)
seasonalityParamsPlot
dev.off()
```

## Visualizing Time Non-Linear Effects and Year Random Effects

```{r}
randomParams <- randomParams %>%
  mutate(timeID=str_match(name,'timeID2(\\d+)')[,2],
         timeID=if_else(is.na(timeID),1L,as.integer(timeID)),
         yearID=str_match(name,'yearID(\\d+)')[,2],
         yearID=if_else(is.na(yearID),1L,as.integer(yearID)))
```

First we visualize the non-linear effects of time (AR1 component)

```{r}
timeParamsPlot <- randomParams %>%
  filter(str_starts(name,'timeID2')) %>%
  mutate(time = make_date(2015+((timeID-1) %/% 12),((timeID-1) %% 12)+1,1)) %>%
  ggplot() +
  geom_line(mapping=aes(x=time,y=mean)) +
  geom_ribbon(mapping=aes(x=time,ymin=`0.025quant`,ymax=`0.975quant`),
              alpha=0.2) +
  scale_x_date(date_breaks = '4 months', date_labels = '%b-%Y', expand = c(0.01,0.01)) +
  labs(x='',
       y='Time Non-Linear Effect (AR1)') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=90,hjust=1),
        panel.grid.minor = element_blank())

timeParamsPlot
```

```{r}
pdf(here(validationFolder,'timeParamsPlot.pdf'), width = 10, height = 6)
timeParamsPlot
dev.off()
```


## Summarizing the Posterior Distribution of Hyperparameters and the Linear Time Trend

The final piece is a table with summary information on the posterior distribution of the hyperparameters and the linear effect of time (which we place here for lack of a better place).

```{r}
hyperParams %>%
  gt() %>%
  cols_label(name='',
             mean='Mean',
             sd='S.D.',
             `0.025quant`='2.5th',
             `0.5quant`='50th',
             `0.975quant`='97.5th') %>%
  tab_spanner(`0.025quant`:`0.975quant`,
              label='Percentile') %>%
  fmt_number(columns = mean:`0.975quant`,
             decimals = 4 )%>%
  gt::gtsave(filename = here(validationFolder,"hyperParamsSummary.png"),zoom=4)
```





