---
title: "Visualizations for journal submission"
author: "Joe A. Wasserman"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
  toc: true
toc_float: true
---
  
```{r setup, include=FALSE}
library(knitr)
library(ragg)
library(here)
library(extrafont)
library(colorspace)
library(tidyverse)
library(ggrepel)
library(ggh4x)
library(scales)
library(patchwork)
library(geofacet)
library(lubridate)

# prompt to register system fonts
# likely to take a long time, but only needs to be done once
# NOTE: requires a different version of Rttf2pt1 to work
# remotes::install_version("Rttf2pt1", version = "1.3.8")
if(is.null(extrafont::fonts())) extrafont::font_import()
loadfonts()

# set default chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  include = FALSE,
  dev = "ragg_png"
)

# set global options
options(
  scipen = 999,
  verbose = FALSE
)

# set directories
here::i_am("R/plotsForSubmission.Rmd")

inDir <- here::here("data", "input")
outDir <- here::here("data", "output")

theme_minimal2 <- theme_minimal() +
  theme(
    line = element_line(color = "grey40", linewidth = .2),
    text = element_text(
      family = "Arial",
      color = "grey10", 
      size = 10
    ),
    rect = element_rect(color = "grey40", fill = "transparent", linewidth = .2)
  )
```

```{r import data}
monthlyEstimates <- read_csv(here::here(outDir, "estimatesMonthly.csv"))
metroRaw <- read_csv(here::here(inDir, "utilities", "FIPSmetroregion4cat.csv"))
# BEARegions <- read_rds(here::here(inDir, "utilities", "stateBEARegionCrosswalk.rds"))
# censusRegion <- read_csv(here::here(inDir, "utilities", "FIPSmetroregion4cat.csv"))
# covidMonthly <- tibble(arrow::read_feather(here::here(outDir,'COVIDMonthlyData.feather')))
```

```{r plot data prep, cache=TRUE}
# Construct data frame of state abbreviations + divisions, plus DC
censusDivision <- tibble(
  state = state.name,
  stateAbbr = state.abb,
  censusDivision = state.division,
  censusRegion = state.region
) %>%
  add_row(
    state = "District of Columbia",
    stateAbbr = "DC",
    censusDivision = "South Atlantic",
    censusRegion = "South"
  ) %>% 
  mutate(
    censusRegion = str_replace(censusRegion, "North Central", "Midwest")
  )

# covidDeaths <- covidMonthly %>% 
#   transmute(
#     monthYear = lubridate::ymd(paste(year, month, "1", sep = "-")),
#     FIPSCode,
#     # COVIDDeaths,
#     imputedCOVIDDeaths
#   )

# filter to pandemic months
plotData <- monthlyEstimates %>%
  filter(monthYear >= date("2020-03-01")) %>%
  arrange(FIPSCode, monthYear) %>% 
  # left_join(covidDeaths, by = c("FIPSCode", "monthYear")) %>% 
  left_join(censusDivision, by = "state") %>%
  mutate(
    excDeaths100k = 100000 * excDeathsMed / pop,
    excDeathsRelative = excDeathsMed / expDeathsMed,
    significant = coalesce(excDeathsLow > 0, FALSE)
  ) %>% 
  group_by(FIPSCode) %>% 
  mutate(popMax = max(pop, na.rm = TRUE)) %>% 
  ungroup()

# color scale for corresponding number of bins
colorscaleGeyser21 <- colorspace::divergingx_hcl(21)

colorscaleRedYellow11 <- colorspace::sequential_hcl(
  palette = "Red-Yellow",
  n = 11L,
  rev = TRUE
)

## Metropolitan status
metro <- metroRaw %>%
  select(
    FIPSCode = fips,
    metroCode = metro
  ) %>%
  mutate(
    FIPSCode = if_else(nchar(FIPSCode) < 5, paste0("0", FIPSCode), as.character(FIPSCode)),
    metroStatus = if_else(
      metroCode %in% c(1, 2),
      "Large Metro",
      "Other"
    ),
    metroStatus2 = case_when(
      metroCode %in% c(1, 2) ~ "Large Metro",
      metroCode %in% c(3) ~ "Medium or Small Metro",
      TRUE ~ "Nonmetro"
    ),
    metroStatus3 = case_when(
      metroCode %in% c(1, 2, 3) ~ "All Metro",
      TRUE ~ "Nonmetro"
    )
  )

```

```{r heatmap data prep, cache=TRUE}
# identify deciles to guide setting cut-points for bins for heatmaps
# plotDataHeatmap %>% filter(excDeathsRelative > 0) %>% pull(excDeathsRelative) %>% quantile(probs = seq(0, 1, .05), type = 8, digits = 4)
# .3, .6, .8, .95
breaksHeatmap <- c(-Inf, 0, .15, .35, .55, 1, Inf)

colorscaleHeatmap <- c("white", colorscaleRedYellow11[c(2, 4, 6, 8, 11)])

plotDataHeatmap <- plotData %>%
  left_join(metro, by = "FIPSCode") %>%
  group_by(FIPSCode) %>%
  mutate(
    excDeathsRelativeNoInf = na_if(excDeathsRelative, Inf),
    excDeathsRelativeNoInf = na_if(excDeathsRelativeNoInf, -Inf),
    monthRank = rank(-excDeathsRelativeNoInf, ties.method = "first", na.last = "keep"),
    monthMax1 = if_else(monthRank == 1L, monthYear, NA_Date_),
    monthMax2 = if_else(monthRank == 2L, monthYear, NA_Date_),
    monthMax3 = if_else(monthRank == 3L, monthYear, NA_Date_)
  ) %>%
  fill(
    monthMax1,
    monthMax2,
    monthMax3,
    .direction = "updown"
  ) %>%
  ungroup() %>%
  arrange(
    # maxMonth
    monthMax1,
    monthMax2,
    monthMax3
  ) %>%
  mutate(
    excDeathsRelativeBinned = cut(
      excDeathsRelative,
      breaks = breaksHeatmap,
      ordered_result = TRUE,
      right = FALSE
    ),
    FIPSPlotOrder = forcats::as_factor(FIPSCode),
    # create 12 evenly-spaced datetimes for nicer plotting
    monthNum = month(monthYear),
    yearStart = floor_date(monthYear, unit = "years"),
    yearEnd = ceiling_date(monthYear, unit = "years"),
    intervalSeconds = int_length(interval(yearStart, yearEnd)) / 12,
    monthYearPlot = yearStart + hours(14) + seconds((monthNum - 1) * intervalSeconds),
    monthTest = monthNum == month(monthYearPlot),
    FIPSlabel = as.character(glue::glue("{countyName}, {stateAbbr}"))
  ) %>% 
  arrange(metroStatus2, FIPSPlotOrder) %>% 
  mutate(
    FIPSPlotOrder = as_factor(FIPSlabel)
  )


# test that dates modified for plotting still fall within same month as in the original data
testthat::expect_true(all(plotDataHeatmap[["monthTest"]]))

FIPSTop50 <- plotDataHeatmap %>% 
  group_by(FIPSCode, metroStatus2) %>% 
  summarize(popMax = max(pop, na.rm = TRUE), .groups = "drop") %>% 
  group_by(metroStatus2) %>% 
  arrange(desc(popMax), .by_group = TRUE) %>% 
  slice(1:50) %>% 
  ungroup()

plotDataHeatmapSubset <- plotDataHeatmap %>% 
  semi_join(FIPSTop50, by = "FIPSCode") %>% 
  filter(metroStatus2 %in% c("Large Metro", "Nonmetro"))
```

# Extreme Excess

```{r extreme excess, include = TRUE, output='asis'}
excessSummary <- plotDataHeatmap %>% 
  group_by(
    state,
    county = countyName,
    FIPSCode,
    censusDivision,
    metroGroup = metroStatus2,
    pop = popMax,
    peakMonth = monthMax1
  ) %>% 
  summarize(
    maxExcess = max(excDeathsRelative),
    monthsOver100Pct = sum(excDeathsRelative > 1),
    cumulativeExcess = 
      sum(excDeathsMed, na.rm = TRUE) / 
      sum(expDeathsMed, na.rm = TRUE),
    .groups = "drop"
  )

excessExtreme <- excessSummary %>% 
  filter(
    monthsOver100Pct > 2 |
      is.na(monthsOver100Pct) |
      is.na(peakMonth) |
      maxExcess > 2 |
      cumulativeExcess > 1
  )

plotExcessExtreme <- plotDataHeatmap %>% 
  semi_join(excessExtreme, by = "FIPSCode")

readr::write_csv(
  excessExtreme,
  here::here(outDir, "countiesWithExcessiveExcess.csv")
)

excessExtreme %>% 
  knitr::kable(digits = 2) %>% 
  kableExtra::kable_styling(
    bootstrap_options = "condensed"
  ) %>% 
  print()
```
```{r extreme excess descriptive plots}
gExcessExtreme <- ggplot(
  plotExcessExtreme,
  aes(x = monthYear)
)

excessExtremeOut <- gExcessExtreme +
  geom_hline(
    yintercept = 0,
    color = "grey40",
    linewidth = .2
  ) +
  ggh4x::stat_difference(
    aes(
      ymin = excDeathsMed, 
      ymax = imputedCOVIDDeaths
    ),
    levels = c("Covid-19 Deaths", "Excess Deaths", NA),
    alpha = .75
  ) +
  geom_line(
    aes(
      y = excDeathsMed,
      color = "Excess Deaths"
    )
  ) +
  geom_line(
    aes(
      y = imputedCOVIDDeaths,
      color = "Covid-19 Deaths"
    )
  ) +
  geom_line(
    aes(
      y = expDeathsMed
    ),
    color = "red",
    linewidth = .3,
    linetype = "dotted"
  ) +
  scale_color_discrete_diverging(
    palette = "Vik",
    rev = TRUE,
    nmax = 10,
    order = c(2, 9)
  ) +
  scale_fill_discrete_diverging(
    palette = "Vik",
    rev = TRUE,
    nmax = 10,
    order = c(4, 7)
  ) +
  scale_x_date(
    breaks = scales::breaks_width(width = "6 months"),
    minor_breaks = scales::breaks_width(width = "3 months"),
    labels = scales::label_date_short(),
    guide = "axis_minor",
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = scales::breaks_extended(n = 3),
    labels = scales::label_comma()
  ) +
  ggh4x::facet_nested_wrap(
    facets = vars(state, metroStatus2, countyName),
    ncol = 11,
    scales = "free_y",
    nest_line = element_line(),
    strip = strip_nested(
      clip = "off",
      by_layer_x = TRUE
    ),
    axes = "all",
    remove_labels = "x"
  ) +
  labs(
    y = "Deaths",
    x = NULL,
    color = NULL,
    fill = NULL
  ) +
  theme_minimal2 +
  theme(
    panel.grid.minor = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.length.x = unit(2, "points"),
    ggh4x.axis.ticks.length.minor = rel(.6),
    axis.text.x = element_blank(),
    legend.position = "bottom",
    legend.justification = "center",
    legend.title.align = .5
  )
```
```{r export extreme excess}
ragg::agg_png(
  filename = here::here("figures", "excessExtreme.png"),
  width = 4000,
  height = 14400,
  res = 300
)
excessExtremeOut
dev.off()
```

# Heatmaps of relative county-month excess mortality as a % (excess / expected)

* Large metro = NCHS categories 1 + 2

* Small/Medium Metro = NCHS category 3

* All Metro = NCH categories 1, 2, + 3

* Non-metro = 4



Largest 50 counties per metro category

Rows (counties) ordered by peak month + second-highest month

```{r heatmap plot, include=TRUE}
# plot waves as line segment annotations
# waveRangePlot <- ggplot(waveRanges) +
#   geom_segment(
#     aes(
#       x = startWave_plot,
#       xend = endWave_plot
#     ),
#     y = 0,
#     yend = 0,
#     color = "grey70"
#   ) +
#   geom_text(
#     aes(
#       x = midWave_plot,
#       y = -.5 - (yjust * 1),
#       label = wave
#     ),
#     vjust = 1.5,
#     color = "grey10", 
#     size = 2.5
#   ) +
#   coord_fixed(2628000 * .8, clip = "off") +
#   theme_void()

# function to plot heatmap and "significance" side by side
plot_heatmap <- function(.x,
                         plot_title = NULL) {
  gHeatmap <- ggplot(
    .x,
    aes(
      y = FIPSPlotOrder,
      x = monthYearPlot,
      fill = excDeathsRelativeBinned,
      width = intervalSeconds
    )
  )
  
  heatmapUnfaceted <- gHeatmap +
    geom_tile(
      color = "transparent"
    ) +
    scale_fill_manual(
      values = colorscaleHeatmap,
      labels = c(
        "≤0%",
        "0-15%",
        "15-30%",
        "30-55%",
        "55-100%",
        ">100%"
      )
    ) +
    scale_x_datetime(
      breaks = scales::breaks_width(width = "6 months", offset = "1 month"),
      minor_breaks = scales::breaks_width(width = "3 months", offset = "1 month"),
      labels = scales::label_date_short(),
      guide = "axis_minor",
      expand = c(0, 0)
    ) +
    scale_y_discrete(
      # labels = NULL,
      limits = rev,
      expand = c(0, 0)
    ) +
    labs(
      title = plot_title,
      subtitle = "Relative Excess Mortality",
      x = NULL,
      y = NULL,
      fill = "Relative Excess Mortality (%)"
    ) +
    coord_fixed(2628000 * 1.3, expand = FALSE) +
    guides(
      fill = guide_legend(
        reverse = FALSE,
        nrow = 1,
        title.position = "top",
        override.aes = list(
          color = "grey40",
          size = .2
        )
      )
    ) +
    theme_minimal2 +
    theme(
      panel.border = element_rect(fill = "transparent"),
      panel.grid = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.length.x = unit(3, "points"),
      ggh4x.axis.ticks.length.minor = rel(.6),
      legend.position = "bottom",
      legend.justification = "center",
      legend.title.align = .5,
      legend.key.size = unit(0.5, "cm")
    )
  
  significantTileUnfaceted <- gHeatmap +
    geom_tile(
      aes(
        fill = significant
      ),
      color = "transparent"
    ) +
    scale_fill_manual(
      values = c("white", "grey70"),
      labels = c(
        "≤95%",
        ">95%"
      )
    ) +
    scale_x_datetime(
      breaks = scales::breaks_width(width = "6 months", offset = "1 month"),
      minor_breaks = scales::breaks_width(width = "3 months", offset = "1 month"),
      labels = scales::label_date_short(),
      guide = "axis_minor",
      expand = c(0, 0)
    ) +
    scale_y_discrete(
      labels = NULL,
      limits = rev,
      expand = c(0, 0)
    ) +
    labs(
      title = NULL,
      subtitle = "Probability of Excess Mortality > 0",
      x = NULL,
      y = NULL,
      fill = "Probability of Excess Mortality > 0"
    ) +
    coord_fixed(2628000 * 1.3, expand = FALSE) +
    guides(
      fill = guide_legend(
        reverse = FALSE,
        nrow = 2,
        title.position = "top",
        override.aes = list(
          color = "grey40",
          size = .2
        )
      )
    ) +
    theme_minimal2 +
    theme(
      panel.border = element_rect(fill = "transparent"),
      panel.grid = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.length.x = unit(3, "points"),
      ggh4x.axis.ticks.length.minor = rel(.6),
      legend.position = "bottom",
      legend.justification = "center",
      legend.title.align = .5,
      legend.key.size = unit(0.5, "cm")
    )
  
  heatmapOutList <- list(heatmapUnfaceted, significantTileUnfaceted)
}

(heatmapOut <- plotDataHeatmapSubset %>% 
    group_by(metroStatus2) %>% 
    group_map(plot_heatmap) %>% 
    map(wrap_plots) %>% 
    map(~ .x[[1]] + .x[[2]]) %>% 
    wrap_plots(
      nrow = 1,
      guides = "collect"
    ) &
    theme(
      legend.position = "bottom",
      legend.justification = "center",
      legend.title.align = .5
    )
)
```
```{r export heatmap}
ragg::agg_png(
  filename = here::here("figures", "heatmap.png"),
  width = 3600,
  height = 2100,
  res = 300
)
heatmapOut
dev.off()
```
```{r appendix heatmaps, eval=FALSE}
# heatmapAppendixList <- plotDataHeatmap %>% 
#   group_by(metroStatus2) %>% 
#   group_map(plot_heatmap) %>% 
#    map(wrap_plots)

gHeatmapAppendix <- ggplot(
  plotDataHeatmap,
  aes(
    y = FIPSPlotOrder,
    x = monthYearPlot,
    fill = excDeathsRelativeBinned,
    width = intervalSeconds
  )
)

heatmapAllCounty <- gHeatmapAppendix +
  geom_tile(
    color = "transparent"
  ) +
  scale_fill_manual(
    values = colorscaleHeatmap,
    labels = c(
      "≤0%",
      "0-15%",
      "15-30%",
      "30-55%",
      "55-100%",
      ">100%"
    )
  ) +
  # ggh4x::facet_nested_wrap(
  #   facets = vars(census_division, metroStatus2),
  #   ncol = 3,
  #   scales = "free_y",
  #   nest_line = element_line(),
  #   # resect = unit(4, "points"),
  #   strip = strip_nested(
  #     clip = "off",
  #     text_x = list(
  #       element_text(size = 9),
  #       element_text()
#     ),
#     by_layer_x = TRUE
#   ),
#   axes = "all",
#   remove_labels = "x"
# ) +
ggh4x::facet_grid2(
  cols = vars(metroStatus2),
  rows = vars(census_division),
  scales = "free_y",
  independent = "y",
  axes = "all",
  remove_labels = "x"
) +
  scale_x_datetime(
    breaks = scales::breaks_width(width = "6 months", offset = "1 month"),
    minor_breaks = scales::breaks_width(width = "3 months", offset = "1 month"),
    labels = scales::label_date_short(),
    guide = "axis_minor",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    # labels = NULL,
    limits = rev,
    expand = c(0, 0)
  ) +
  labs(
    title = NULL,
    subtitle = "Relative Excess Mortality",
    x = NULL,
    y = NULL,
    fill = "Relative Excess Mortality (%)"
  ) +
  # coord_fixed(2628000 * 1.3, expand = FALSE) +
  guides(
    fill = guide_legend(
      reverse = FALSE,
      nrow = 1,
      title.position = "top",
      override.aes = list(
        color = "grey40",
        size = .2
      )
    )
  ) +
  theme_minimal2 +
  theme(
    panel.border = element_rect(fill = "transparent"),
    panel.grid = element_blank(),
    axis.ticks.x = element_line(),
    axis.ticks.length.x = unit(3, "points"),
    ggh4x.axis.ticks.length.minor = rel(.6),
    legend.position = "bottom",
    legend.justification = "center",
    legend.title.align = .5,
    legend.key.size = unit(0.5, "cm")
  )

heatmapAllCounty
```
```{r export appendix heatmaps, eval=FALSE}
# saveHeatmap <- function(plot,
#                         plot_name = "",
#                         nrows = 50) {
#   ragg::agg_png(
#     filename = here::here("figures", paste0("heatmap", plot_name, ".png")),
#     width = 1800,
#     height = nrows * 35 + 200,
#     res = 300
#   )
#   plot
#   dev.off()
# }
# 
# pmap(
#   .l = list(plot = heatmapAppendixList,
#   plot_name = c("Large", "MedSm", "Non"),
#   nrows = c(434, 728, 1965)),
#   .f = ~ saveHeatmap(plot = ..1, plot_name = ..2, nrows = ..3)
# )

```

# Geofacet relative excess line graphs

Note: Line graphs only depict counties with > 20k population


```{r geofacet data prep, cache=TRUE}
plotDataGeofacet <- mutate(
  plotDataHeatmap, 
  plotGroup = "State Total"
)  %>% 
  group_by(
    stateFIPS,
    state,
    monthYear,
    plotGroup
  ) %>% 
  summarize(
    across(
      c(
        expDeathsLow:imputedCOVIDDeaths,
        deaths,pop,
        imputedDeaths,
        excDeathsMed:excDeathsUp
      ),
      sum,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>% 
  mutate(
    excDeathsRelative = excDeathsMed / expDeathsMed
  ) %>% 
  bind_rows(
    mutate(plotDataHeatmap, plotGroup = "Individual Counties")
  ) %>% 
  # NA pop in filter captures state totals
  filter(popMax > 15000 | is.na(popMax))

# grid of states for geofacet plot
us_state_grid_custom <- data.frame(
  row = c(1, 1, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 4, 4, 4, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 6, 6, 6, 6, 6, 6, 6, 6, 6, 7, 7, 7, 7, 7, 7, 8, 8),
  col = c(11, 1, 9, 10, 11, 10, 6, 5, 4, 3, 2, 1, 9, 8, 11, 6, 5, 4, 3, 2, 1, 7, 9, 8, 9, 7, 6, 5, 4, 3, 2, 1, 10, 8, 6, 5, 4, 3, 2, 1, 7, 9, 8, 6, 5, 4, 3, 8, 7, 1, 7),
  code = c("ME", "AK", "VT", "NH", "MA", "CT", "MI", "MN", "SD", "ND", "MT", "WA", "NY", "PA", "RI", "WI", "IA", "NE", "WY", "ID", "OR", "IN", "NJ", "WV", "MD", "OH", "IL", "MO", "KS", "CO", "UT", "NV", "DE", "DC", "KY", "AR", "OK", "NM", "AZ", "CA", "TN", "VA", "NC", "AL", "MS", "LA", "TX", "SC", "GA", "HI", "FL"),
  name = c("Maine", "Alaska", "Vermont", "New Hampshire", "Massachusetts", "Connecticut", "Michigan", "Minnesota", "South Dakota", "North Dakota", "Montana", "Washington", "New York", "Pennsylvania", "Rhode Island", "Wisconsin", "Iowa", "Nebraska", "Wyoming", "Idaho", "Oregon", "Indiana", "New Jersey", "West Virginia", "Maryland", "Ohio", "Illinois", "Missouri", "Kansas", "Colorado", "Utah", "Nevada", "Delaware", "District of Columbia", "Kentucky", "Arkansas", "Oklahoma", "New Mexico", "Arizona", "California", "Tennessee", "Virginia", "North Carolina", "Alabama", "Mississippi", "Louisiana", "Texas", "South Carolina", "Georgia", "Hawaii", "Florida"),
  stringsAsFactors = FALSE
)
#   row = c(1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 5, 5, 5, 5, 5, 5, 5, 6, 6, 6, 6, 6, 6, 7, 7, 7),
#   col = c(12, 11, 11, 12, 4, 10, 8, 5, 3, 6, 2, 1, 11, 12, 4, 10, 9, 6, 7, 5, 3, 8, 2, 1, 2, 10, 9, 4, 5, 7, 3, 8, 1, 6, 11, 7, 9, 3, 6, 4, 5, 8, 11, 6, 7, 8, 5, 4, 2, 9, 1),
#   code = c("ME", "NH", "VT", "MA", "ND", "NY", "MI", "MN", "MT", "WI", "ID", "WA", "CT", "RI", "SD", "NJ", "PA", "IL", "IN", "IA", "NE", "OH", "WY", "OR", "NV", "MD", "VA", "CO", "KS", "KY", "UT", "WV", "CA", "MO", "DE", "TN", "NC", "AZ", "AR", "NM", "OK", "SC", "DC", "MS", "AL", "GA", "LA", "TX", "AK", "FL", "HI"),
#   name = c("Maine", "New Hampshire", "Vermont", "Massachusetts", "North Dakota", "New York", "Michigan", "Minnesota", "Montana", "Wisconsin", "Idaho", "Washington", "Connecticut", "Rhode Island", "South Dakota", "New Jersey", "Pennsylvania", "Illinois", "Indiana", "Iowa", "Nebraska", "Ohio", "Wyoming", "Oregon", "Nevada", "Maryland", "Virginia", "Colorado", "Kansas", "Kentucky", "Utah", "West Virginia", "California", "Missouri", "Delaware", "Tennessee", "North Carolina", "Arizona", "Arkansas", "New Mexico", "Oklahoma", "South Carolina", "District of Columbia", "Mississippi", "Alabama", "Georgia", "Louisiana", "Texas", "Alaska", "Florida", "Hawaii"),
#   stringsAsFactors = FALSE
# )

```

```{r geofacet plot, include=TRUE}
gGeofacet <- ggplot(
  plotDataGeofacet,
  aes(
    x = monthYear,
    y = excDeathsRelative * 100,
    # linetype = metroStatus,
    group = countyName,
    color = plotGroup,
    alpha = plotGroup
  )
)

(geofacetOut <- gGeofacet +
    geom_hline(
      yintercept = 0,
      color = "grey40",
      linewidth = .2
    ) +
    geom_line(
      # color = "grey10",
      linewidth = .3
    ) +
    scale_alpha_discrete(
      range = c(.21, 1)
    ) +
    scale_color_discrete_diverging(
      palette = "Vik",
      rev = TRUE,
      nmax = 10,
      order = c(4, 10)
    ) +
    scale_x_date(
      # expand = c(0, 0),
      breaks = scales::breaks_width(width = "6 months"),
      minor_breaks = scales::breaks_width(width = "3 months"),
      labels = scales::label_date_short(),
      guide = "axis_minor"
    ) +
    scale_y_continuous(
      # expand = c(0, 0),
      oob = scales::oob_keep,
      breaks = c(seq.int(-100, 200, 50)),
      minor_breaks = c(seq.int(-75, 175, 50)),
      guide = "axis_minor"
    ) +
    coord_cartesian(
      ylim = c(-50, 200),
      expand = FALSE,
      clip = "off"
    ) +
    facet_geo(
      facets = ~ state,
      grid = us_state_grid_custom
    ) +
    labs(
      y = "Relative Excess Mortality (%)",
      x = NULL,
      color = NULL,
      linetype = NULL,
      alpha = NULL
    ) +
    guides(
      color = "legend",
      alpha = "none"
    ) +
    theme_minimal2 +
    theme(
      text = element_text(size = 9.8),
      panel.border = element_rect(fill = "transparent"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(),
      axis.ticks = element_line(),
      axis.ticks.length = unit(3.5, "points"),
      ggh4x.axis.ticks.length.minor = rel(.61),
      legend.position = "bottom",
      legend.justification = "center"
    )
)
```
```{r export geofacet}
ragg::agg_png(
  filename = here::here("figures", "geofacet.png"),
  width = 3600,
  height = 2500,
  res = 300
)
geofacetOut
dev.off()
```

# Facet line graphs of excess vs COVID deaths
```{r excess v covid data prep}
plotDataExcessVSCovid <- plotDataHeatmap %>% 
  group_by(monthYear, metroStatus2, census_division) %>% 
  summarize(
    across(
      c(
        expDeathsLow:imputedCOVIDDeaths,
        deaths:pop,
        imputedDeaths,
        excDeathsMed:excDeathsUp
      ),
      sum,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>% 
  mutate(
    excDeathsRelative = excDeathsMed / expDeathsMed
  )

plotDataExcessVSCovidSubset <- plotDataHeatmap %>% 
  filter(pop > 20000) %>% 
  group_by(monthYear, metroStatus2, census_division) %>% 
  summarize(
    across(
      c(
        expDeathsLow:imputedCOVIDDeaths,
        deaths:pop,
        imputedDeaths,
        excDeathsMed:excDeathsUp
      ),
      sum,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>% 
  mutate(
    excDeathsRelative = excDeathsMed / expDeathsMed
  )
```
```{r excess v covid plot, include=TRUE}
gExcessVSCovid <- ggplot(
  plotDataExcessVSCovid,
  aes(x = monthYear)
)

(
  differenceExcessVSCovidOut <- gExcessVSCovid +
    geom_hline(
      yintercept = 0,
      color = "grey40",
      linewidth = .2
    ) +
    ggh4x::stat_difference(
      aes(
        ymin = excDeathsMed, 
        ymax = imputedCOVIDDeaths
      ),
      levels = c("Covid-19 Deaths", "Excess Deaths", NA),
      alpha = .75
    ) +
    geom_line(
      aes(
        y = excDeathsMed,
        color = "Excess Deaths"
      ),
      linewidth = .35
    ) +
    geom_line(
      aes(
        y = imputedCOVIDDeaths,
        color = "Covid-19 Deaths"
      ),
      linewidth = .35
    ) +
    scale_color_discrete_diverging(
      palette = "Vik",
      rev = TRUE,
      nmax = 10,
      order = c(2, 9)
    ) +
    scale_fill_discrete_diverging(
      palette = "Vik",
      rev = TRUE,
      nmax = 10,
      order = c(4, 7)
    ) +
    scale_x_date(
      breaks = scales::breaks_width(width = "6 months"),
      minor_breaks = scales::breaks_width(width = "3 months"),
      labels = scales::label_date_short(),
      guide = "axis_minor",
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      breaks = scales::breaks_extended(n = 5),
      labels = scales::label_comma()
    ) +
    ggh4x::facet_nested_wrap(
      facets = vars(census_division, metroStatus2),
      ncol = 3,
      scales = "free_y",
      nest_line = element_line(),
      # resect = unit(4, "points"),
      strip = strip_nested(
        clip = "off",
        text_x = list(
          element_text(size = 9),
          element_text()
        ),
        by_layer_x = TRUE
      ),
      axes = "all",
      remove_labels = "x"
    ) +
    # ggh4x::facet_grid2(
    #   rows = vars(censusDivision),
    #   cols = vars(metroStatus2),
    #   scales = "free_y",
    #   independent = "y",
    #   strip = strip_nested(clip = "off"),
    #   axes = "all",
    #   remove_labels = "x"
    # ) +
    labs(
      y = "Deaths",
      x = NULL,
      color = NULL,
      fill = NULL
    ) +
    theme_minimal2 +
    theme(
      panel.grid.minor = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.length.x = unit(3, "points"),
      ggh4x.axis.ticks.length.minor = rel(.6),
      legend.position = "bottom",
      legend.justification = "center",
      legend.title.align = .5,
      strip.text.y = element_text(
        angle = 0,
        hjust = 0
      )
    )
)

```
```{r export excess v covid}
ragg::agg_png(
  filename = here::here("figures", "differenceExcessVsCovid.png"),
  width = 2100,
  height = 3300,
  res = 300
)
differenceExcessVSCovidOut
dev.off()
```
```{r excess v covid plot subset comparison}
# This plot excludes small counties (< 30k population) as a check to see whether
# regions with COVID > excess are a data artifact

gExcessVSCovidSubset <- ggplot(
  plotDataExcessVSCovidSubset,
  aes(x = monthYear)
)

(
  differenceExcessVSCovidSubset <- gExcessVSCovidSubset +
    geom_hline(
      yintercept = 0,
      color = "grey40",
      linewidth = .2
    ) +
    geom_line(
      aes(
        y = excDeathsMed,
        color = "Excess Deaths"
      )
    ) +
    geom_line(
      aes(
        y = imputedCOVIDDeaths,
        color = "Covid-19 Deaths"
      )
    ) +
    ggh4x::stat_difference(
      aes(
        ymin = excDeathsMed, 
        ymax = imputedCOVIDDeaths
      ),
      levels = c("Covid-19 Deaths", "Excess Deaths", NA),
      alpha = .75
    ) +
    scale_color_discrete_diverging(
      palette = "Vik",
      rev = TRUE,
      nmax = 10,
      order = c(2, 9)
    ) +
    scale_fill_discrete_diverging(
      palette = "Vik",
      rev = TRUE,
      nmax = 10,
      order = c(4, 7)
    ) +
    scale_x_date(
      breaks = scales::breaks_width(width = "6 months"),
      minor_breaks = scales::breaks_width(width = "3 months"),
      labels = scales::label_date_short(),
      guide = "axis_minor",
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      breaks = scales::breaks_extended(n = 5),
      labels = scales::label_comma()
    ) +
    ggh4x::facet_nested_wrap(
      facets = vars(census_division, metroStatus2),
      ncol = 3,
      scales = "free_y",
      nest_line = element_line(),
      # resect = unit(4, "points"),
      strip = strip_nested(
        clip = "off",
        text_x = list(
          element_text(size = 9),
          element_text()
        ),
        by_layer_x = TRUE
      ),
      axes = "all",
      remove_labels = "x"
    ) +
    # ggh4x::facet_grid2(
    #   rows = vars(censusDivision),
    #   cols = vars(metroStatus2),
    #   scales = "free_y",
    #   independent = "y",
    #   strip = strip_nested(clip = "off"),
    #   axes = "all",
    #   remove_labels = "x"
    # ) +
    labs(
      y = "Deaths",
      x = NULL,
      color = NULL,
      fill = NULL,
      title = "EXCLUDES COUNTIES < 20K -- NOT FOR PUBLICATION"
    ) +
    theme_minimal2 +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.ticks.x = element_line(),
      axis.ticks.length.x = unit(3, "points"),
      ggh4x.axis.ticks.length.minor = rel(.6),
      legend.position = "bottom",
      legend.justification = "center",
      legend.title.align = .5,
      strip.text.y = element_text(
        angle = 0,
        hjust = 0
      )
    )
)

```
```{r export excess v covid subset comparison, include=FALSE}
ragg::agg_png(
  filename = here::here("figures", "differenceExcessVSCovidSubsetCheck.png"),
  width = 2100,
  height = 3300,
  res = 300
)
differenceExcessVSCovidSubset
dev.off()
```

# "Cumulative" Excess Deaths
```{r cumulative excess data prep}
plotCumulative <- plotDataHeatmap %>% 
  arrange(censusRegion, metroStatus2, monthYear) %>% 
  group_by(censusRegion, metroStatus2, monthYear) %>% 
  summarize(
    across(
      c(
        expDeathsLow:imputedCOVIDDeaths,
        deaths:pop,
        imputedDeaths,
        excDeathsMed:excDeathsUp
      ),
      sum,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) %>% 
  group_by(censusRegion, metroStatus2) %>% 
  mutate(
    across(
      c(expDeathsLow:imputedCOVIDDeaths, imputedDeaths:excDeathsUp),
      ~ with_order(monthYear, cumsum, .x),
      .names = "{.col}Cumulative"
    ),
    excDeathsRelativeCumulative = excDeathsMedCumulative / expDeathsMedCumulative,
    significantCumulative = excDeathsLow > 0,
    metroAbbr = str_replace_all(
      metroStatus2,
      c(
        "Large" = "Lrg",
        "Medium or Small" = "Med/Sm"
      )
    ),
    labelMargin = as.character(
      glue::glue(
        "{censusRegion} – {metroStatus2}"
      )
    ),
    labelPoint = as.character(
      glue::glue(
        "{censusRegion}\n{metroAbbr}"
      )
    )
  ) %>% 
  ungroup() %>% 
  mutate(
    labelMargin = if_else(
      monthYear == ymd("2022-02-01"),
      labelMargin,
      NA_character_
    ),
    labelPointTop = if_else(
      censusRegion == "Northeast" &
        metroStatus2 == "Large Metro" &
        monthYear == ymd("2020-04-01"),
      labelPoint,
      NA_character_
    ),
    labelPointBottom = if_else(
      (
        censusRegion == "North Central" &
          metroStatus2 == "Large Metro" &
          monthYear == ymd("2020-05-01")
      ) |
        (
          censusRegion == "North Central" &
            metroStatus2 == "Large Metro" &
            monthYear %in% c(
              ymd("2020-05-01")
            )
        ) |
        (
          censusRegion == "Northeast" &
            metroStatus2 == "Medium or Small Metro" &
            monthYear %in% c(
              ymd("2020-05-01")
            )
        ),
      labelPoint,
      NA_character_
    ),
    pointForLabel = !(is.na(labelPointTop) & is.na(labelPointBottom))
  )
```
```{r cumulative excess plot}
gExcessCumulative <- ggplot(
  plotCumulative,
  aes(
    x = monthYear,
    y = excDeathsRelativeCumulative,
    color = censusRegion,
    linetype = metroStatus2
  )
)

excessCumulativeShared <- gExcessCumulative +
  geom_line(
    linewidth = .375
  ) +
  # geom_point(
  #   inherit.aes = FALSE,
  #   x = ymd("2020-03-01"),
  #   y = .3,
  #   shape = "/",
  #   show.legend = FALSE,
  #   color = "grey30"
  # ) +
  # geom_point(
  #   data = filter(plotCumulative, pointForLabel),
  #   aes(fill = censusRegion),
#   size = .5
# ) +
scale_color_discrete_qualitative(palette = "Harmonic") +
  scale_linetype_manual(
    values = c(
      "solid",
      "32",
      "12"
    )
  ) +
  scale_x_date(
    breaks = scales::breaks_width(width = "3 months"),
    minor_breaks = scales::breaks_width(width = "1 months"),
    labels = scales::label_date_short(),
    guide = "axis_minor",
    oob = scales::oob_keep
  ) +
  labs(
    y = "Relative Excess Mortality (%)",
    x = NULL,
    color = "group",
    linetype = "group"
  ) +
  guides(
    color = "none",
    linetype = "none",
    fill = "none",
    label = "none"
  ) +
  theme_minimal2 +
  theme(
    text = element_text(size = 11.5),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(),
    axis.ticks = element_line(),
    axis.ticks.length = unit(3.5, "points"),
    ggh4x.axis.ticks.length.minor = rel(.61)
  )

excessCumulativeBottom <- excessCumulativeShared +
  geom_point(
    data = tibble(x = c(ymd("2020-03-06"), ymd("2020-09-29"))),
    aes(x = x),
    y = .3,
    color = colorspace::qualitative_hcl(palette = "Harmonic", n = 4)[[2]],
    shape = "-",
    size = 4,
    inherit.aes = FALSE
  ) +
  # ggrepel::geom_label_repel(
  #     aes(label = labelPointBottom),
  #     size = 3,
  #     segment.size = .2,
  #     label.size = 0,
  #     min.segment.length = 0,
  #     force_pull = .5,
  #     seed = 20221030,
  #     show.legend = FALSE
  #   ) +
  scale_y_continuous(
    breaks = seq(0, .3, .1),
    minor_breaks = seq(.05, .25, .1),
    oob = scales::oob_keep,
    labels = ~ .x * 100
  ) +
  coord_cartesian(
    ylim = c(0, .3003),
    expand = FALSE,
    clip = "on"
  ) +
  labs(
    y = "Relative Excess Mortality (%)",
    x = NULL,
    color = NULL,
    linetype = NULL
  ) +
  theme(
    plot.margin = unit(c(0, 0, 5.5, 5.5), "points")
  )

excessCumulativeTop <- excessCumulativeShared +
  geom_point(
    data = tibble(x = c(ymd("2020-03-06"), ymd("2020-09-28"))),
    aes(x = x),
    y = .3005,
    color = colorspace::qualitative_hcl(palette = "Harmonic", n = 4)[[2]],
    shape = "-",
    size = 4,
    inherit.aes = FALSE
  ) +
  # ggrepel::geom_label_repel(
  #   aes(label = labelPointTop),
  #   size = 3,
  #   segment.size = .2,
  #   label.size = 0,
  #   min.segment.length = 0,
  #   force_pull = .5,
  #   seed = 20221030,
  #   show.legend = FALSE
  # ) +
  scale_y_continuous(
    breaks = seq(.4, 1, .1),
    minor_breaks = seq(.35, .95, .1),
    oob = scales::oob_keep,
    labels = ~ .x * 100
  ) +
  coord_cartesian(
    ylim = c(.2995, 1),
    expand = FALSE,
    clip = "on"
  ) +
  labs(
    y = NULL,
    x = NULL,
    color = NULL,
    linetype = NULL
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    plot.margin = unit(c(5.5, 0, 0, 5.5), "points")
  )

excessCumulativeLabelMargin <- gExcessCumulative +
  geom_text_repel(
    aes(label = labelMargin),
    size = 3.5,
    segment.size = .25,
    segment.linetype = "16",
    # segment.color = "grey60",
    direction = "y",
    hjust = 0,
    force_pull = .2,
    xlim = c(ymd("2022-02-02"), Inf),
    seed = 20221030,
    show.legend = FALSE
  ) +
  scale_y_continuous(
    breaks = scales::breaks_extended(4),
    oob = scales::oob_keep
  ) +
  coord_cartesian(
    ylim = c(0, .3005),
    xlim = c(ymd("2022-02-01"), ymd("2022-02-10")),
    expand = FALSE,
    clip = "off"
  ) +
  scale_color_discrete_qualitative(palette = "Harmonic") +
  scale_x_date(
    # oob = scales::oob_keep
  ) +
  labs(
    y = NULL,
    x = NULL,
    color = NULL
  ) +
  theme_void() +
  theme(
    plot.margin = unit(c(0, 0, 5.5, 0), "points")
  )

legendDF <- crossing(
  censusDivision = c("Midwest", "Northeast", "South", "West"),
  metroStatus2 = c("Large Metro", "Medium or Small Metro", "Nonmetro")
) %>%
  mutate(
    y = as.integer(as_factor(metroStatus2)),
    x = as.integer(as_factor(censusDivision)) * 3.25 - 2.25
  )

gLegend <- ggplot(
  legendDF,
  aes(
    x = x - .5,
    xend = x + .5,
    y = y + .5,
    yend = y - .5,
    color = censusDivision,
    linetype = metroStatus2
  )
) +
  geom_segment() +
  geom_label(
    aes(
      x = x,
      label = stringr::str_wrap(censusDivision, width = 12)
    ),
    y = 0,
    hjust = .5,
    size = 3.5,
    label.size = 0
  ) +
  geom_label(
    aes(
      y = y,
      label = stringr::str_wrap(metroStatus2, width = 12)
    ),
    x = 12,
    color = "grey10",
    hjust = 0,
    vjust = .5,
    size = 3.5,
    label.size = 0
  ) +
  scale_color_discrete_qualitative(palette = "Harmonic") +
  scale_linetype_manual(
    values = c(
      "solid",
      "32",
      "12"
    )
  ) +
  scale_y_continuous(
    trans = "reverse"
  ) +
  coord_cartesian(
    clip = "off",
    xlim = c(NA, 13)
  ) +
  guides(
    color = "none",
    linetype = "none",
    label = "none"
  ) +
  theme_void() +
  theme(
    plot.margin = unit(c(30, 40, 0, 1), "points"),
    panel.background = element_rect(fill = "white", color = "white"),
    plot.background = element_rect(fill = "white", color = "white")
  )

layout <- c(
  # top
  area(t = 1, l = 1, b = 3, r = 8),
  # bottom
  area(t = 4, l = 1, b = 10, r = 8),
  # margin labels
  area(t = 4, l = 9, b = 10, r = 11),
  # legend
  area(t = 2, l = 4, b = 3, r = 7)
)

(
  excessCumulativeOut <- excessCumulativeTop +
    excessCumulativeBottom +
    excessCumulativeLabelMargin +
    gLegend +
    plot_layout(design = layout)
)
```
```{r export cumulative excess, include=FALSE}
ragg::agg_png(
  filename = here::here("figures", "excessCumulative.png"),
  width = 3600,
  height = 2100,
  res = 300
)
excessCumulativeOut
dev.off()
```
